---
title: "CLABSI SPC Report"
author: "Dr. Shewart"
date: "January 19, 2017"
output: html_document
---

```{r setup, include=FALSE}
# Do not show R code or messages in output
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Since the output will go to the same directory as the code, save it to your home directory instead
setwd("~/")


#### load R packages ####
library(tidyverse)
library(qicharts)
library(gridExtra)
library(ggExtra)
library(forecast)


#### Load data from shared folder ####

#### !!! USE YOUR USERID !!! ####
CLABSI = read_csv("C:\\Users\\dbarr1\\Downloads\\CLABSI.csv")

CLABSI$Month = as.Date(CLABSI$Month)

# Create a time series object of the Rate variable
CLABSI_ts = ts(CLABSI$Rate, start = c(2006, 10), frequency = 12)

# Create a cumulative sums by year data frame
CLABSI_accum = 
    CLABSI %>%
    group_by(FY) %>% 
    arrange(Month) %>% 
    mutate(cuml_linedays = cumsum(CL_Days),
           cuml_infections = cumsum(CLABSI)) 

# Subset data to start at FY11
CLABSI_2010 = filter(CLABSI, Month >= "2010-10-01")
```


## CLABSI EDA

**Did you know** that the mean ($\bar{x}$) <s>count</s> rate is `r mean(CLABSI$Rate)`?  

Note you need *two* spaces at the end of each line to make a line break.  

OK, now gimme some plots.  

Assessing the trend, first version.  

```{r trend1}
# Line plot with loess smoother for assessing trend
p1 = ggplot(CLABSI, aes(x = Month, y = Rate)) + 
  geom_smooth() + 
  geom_line() 

# Histogram with density overlay
p2 = ggplot(CLABSI, aes(Rate)) + 
  geom_histogram(aes(y = ..density..), 
    binwidth = 0.5, color="gray70") +
  geom_density(color="blue")

# Put both into the same plot
grid.arrange(p1, p2, widths = c(0.65, 0.35))
```

Here's another way to look at it.  

```{r trend2}
# Another way to view series, trend, and ditribution together
ggMarginal(p1, margins="y", type = "histogram", binwidth=0.5)
```


### Seasonplot 

```{r seasonplot, fig.width=10, fig.height=4}
ggseasonplot(CLABSI_ts)
```

### Monthplot

```{r monthplot, fig.width=10, fig.height=4}
ggmonthplot(CLABSI_ts)
```

### Accumulation plot

```{r accumplot}
# Plot accumultation curves of CLABSIs by CL days by FY
ggplot(CLABSI_accum, aes(x = cuml_linedays, y = cuml_infections, group = as.factor(FY))) +
    geom_path(aes(color = as.factor(FY)), size = 1) +
    geom_point(aes(color = as.factor(FY)))+
    scale_y_continuous(name = "Cummulative CLABSIs", breaks = seq(0,120,10)) +
    scale_x_continuous(name = "Cumulative Central Line Days", breaks = seq(0,40000,5000)) +
    scale_colour_brewer(type = "div", palette = "Spectral") +
    guides(color = guide_legend(title = "Fiscal\nYear")) +
    ggtitle("CLABSIs vesus Central Line Days by Year")
```

## Control Chart

```{r uchart, fig.width=10, fig.cap="Limits calculated FY11 to present because I checked the assumptions and hey, that's how it goes."}
# Create an empty R object
notes = NA

# Create an annotation at point #48
notes[48] = "Something rad happened" 

qic(y = CLABSI, 
    n = CL_Days, 
    x = Month,
    multiply = 1000,
    breaks = 48,
    data = CLABSI, 
    chart = "u",
    main = "CLABSI Rate per 1000 Central Line Days (u-chart)",
    ylab = "CLABSI Rate",
    xlab = "Month")
```

<br>  

### Oh, here's some data in an interactive table

```{r interactivetable}
# Add a character variable for month
CLABSI$MoY = lubridate::month(CLABSI$Month, label = TRUE, abbr = FALSE)

# Make an interactive table inside the html
DT::datatable(CLABSI, filter = "top")
```

<hr>  

- This is the end.  

<hr>

- Wait, no it's not. Here's some info about this R session:

```{r sessioninfo}
sessionInfo()
```

*~ The end ~ *
